Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate

### 2018-03-19 Paper/main.tex
Beginning to try to print a barplot histogram. Why does the Julia code take so long to run?

\begin{abstract}
Recently, there has been a growing concern in academia, industrial research labs and the mainstream commercial media about the phenomenon of machine bias, where trained statistical models unbeknownst to their creators become vessels of bias and prejudices. A large number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two Asian people and the infamous case of Google photos' classifying black people as gorillas. Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of \emph{gender} bias in AI. 
In this paper, we construct sentences with gender neutral pronouns in different languages 
and translate these sentences into English using the Google Translate API. 
The resulting English sentences are constructed either with male, female or neutral pronouns, which allows us to collect statistics about the frequency of male defaults and evaluate which factors can influence the gender Google Translate 
API computes for each sentence. We show that not only there is a tendency towards translating pronouns to
masculine, but also that this tendency is exaggerated for sentences about male dominated fields such as Computer Science. On the other hand, sentences about artistic fields are less likely to yield male defaults. This possibly reflects the implicit assumptions our society has about stereotypical gender roles, which suggests that without proper care statistical machine translation can become hostage to the prejudices of the people behind their training datasets -- ourselves.
\end{abstract}

@@

@@

% PEDRO: Numbers don't match, but shouldn't. One is a average over all translations and the one below is an average over the average:
These values contrast with LBS' report of gender participation. Running a t-test on the percentage of female translations over all translations for every occupation against the percentage of female participation in an occupation (surrogating with the LBS class value for the profession, in case it was missing), we see a mean of $11.11\%$ gender participation in the translations and a mean of $35.99\%$ in the LBS report. The variance reported for the translation results is also lower, at $0.0145145752$ in contrast with the report's $0.0666766998$. Altogether, the Pearson correlation between the two values is $0.3035612739$, with the p-values for one and two tail both being lower than the alpha value chosen ($\alpha = 0.05$). This result enforces the fact that the translation exhibits a tendency towards male defaults, even when the female participation in that job is higher than expected.

@@

@@

\section{Conclusions}
In this paper, we have provided evidence that statistical translation tools such as Google Translate exhibit implicit gender biases and a tendency towards male defaults, which possibly stem from the real world data used to train them. As a result, we suggest that such tools can be probed to yield insights about stereotypical gender roles in our society. By translating sentences from gender neutral languages such as Hungarian and Chinese into English, we were able to collect statistics about the asymmetry between female and male pronominal genders in the translation process. Because Google Translate typically uses English as a \emph{lingua franca}
% PEDRO: reference?
 to translate between other languages, our findings about gender bias in professional occupations possibly generalize to most translations from gender neutral idioms in the same tool, although we have not tested this hypothesis yet. Our results further show that although male pronouns are typical, the proportion of female pronouns can vary significantly according either to the adjectives used (\say{shy}/\say{happy} vs \say{brave}/\say{guilty}) or the category of professional occupations (artistic jobs are far more likely to be translated with female pronouns than computer-related ones). We have also shown that different languages are differently biased towards male defaults, with Hungarian exhibiting a better balance between male and female pronouns than, say, Chinese. Some languages such as Yoruba and Basque have been found to translate sentences with gender neutral pronouns very often, although this is the exception rather than the rule.

We think this work can shed further light on what we believe is a pressing ethical difficulty arising from modern statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potentially harmful effects of the pressing modern concerns of machine bias.

>>>>

\begin{abstract}
Recently there has been a growing concern in academia, industrial research labs and the mainstream commercial media about the phenomenon dubbed as \emph{machine bias}, where trained statistical models -- unbeknownst to their creators -- grow to reflect controversial societal asymmetries, such as gender or racial bias. A significant number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two distinct Asian people and the now infamous case of Google photos' mistakenly classifying people of color as gorillas. Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of gender bias in AI.

In this paper, we obtained a comprehensive list of job positions from the U.S. Bureau of Labor Statistics and used it to build sentences in the format of \say{He/She is an Engineer} in 11 different gender neutral languages such as Hungarian, Chinese, Yoruba, etc. We translate these sentences into English using the Google Translate API, and collect statistics about the frequency of female, male and gender-neutral pronouns in the translated output. We then show that Google Translate exhibits a strong tendency towards male defaults, which is exaggerated for fields suggested to be troubled with gender stereotypes such as STEM (Science, Technology, Engineering and Mathematics) jobs. We ran these statistics against BLS' data for the frequency of female participation in each job position, in which we show that Google Translate fails to reproduce the real-world distribution of female workers. In summary, we provide evidence that even if one does not expect in principle a 50:50 pronominal gender distribution, Google Translate yields male defaults much more frequently than what would be expected from demographic data alone.

We believe that our study can shed further light on the phenomenon of machine bias and are hopeful that it will ignite a debate about the need for augmented current statistical translation tools with debiasing techniques -- which can already be found in the scientific literature.
\end{abstract}

@@

\section{Assumptions and Preliminaries}

@@

% PEDRO: Numbers don't match, but shouldn't. One is a average over all translations and the one below is an average over the average:
These values contrast with LBS' report of gender participation, which will be discussed in more detail in Section \ref{sec:comparison-women-participation}.

@@

\section{Comparison with women participation data across job positions}

A sensible objection to the conclusions we draw from our study is that the perceived gender bias in Google Translate results stems from the fact that (possibly) female participation in some job positions is itself low. We must account for the possibility that the statistics of gender pronouns in Google Translate outputs merely reflects a real-world asymmetry in the workplace. In this context, the argument in favor of a critical revision of statistic translation algorithms weakens considerably, and possibly shifts the blame away from these tools.

The U.S. Bureau of Labor Statistics data summarized in Table \ref{tab:occupations} contains statistics about the percentage of women participation in each occupation category. These data is also available for each individual occupation, which allows us to compute the frequency of women participation for each 12-quantile. We carried the same computation in the context of frequencies of translated female pronouns, and the resulting histograms are plotted side-by-side in Figure \ref{fig:histogram-compare-gt-real}. The data shows us that Google Translate outputs fail to follow the real-world distribution of female workers across a comprehensive set of job positions. The distribution of translated female pronouns is consistently inversely distributed, with female pronouns accumulating in the first 12-quantile. By contrast, BLS data shows that female participation peaks in the fourth 12-quantile and remains significant throughout the next ones.

% Histogram with the distirbution of male prononuns for each occupation category
\begin{figure}[H]
	\centering
	\includegraphics[width=10cm]{pictures/histogram-compare-gt-real}
	\caption{Women participation ($\%$) data obtained from the U.S. Bureau of Labor Statistics allows us to assess whether the Google Translate bias towards male defaults is at least to some extent explained by small frequencies of female workers in some job positions. Our data does not make a very good case for that hypothesis: the total frequency of translated female pronouns (in blue) for each 12-quantile (which is consistently inversely distributed) does not seem to respond to the higher proportion of female workers (in red) in the last quantiles. To remove doubt we have also plotted in yellow the frequency of translated female pronouns among male and female pronouns (i.e., excluding eventual translation errors and gender-neutral prononuns).}
	\label{fig:histogram-compare-gt-real}
\end{figure}

Running a t-test on the percentage of female translations over all translations for every occupation against the percentage of female participation in an occupation (surrogating with the LBS class value for the profession, in case it was missing), we see a mean of $11.11\%$ gender participation in the translations and a mean of $35.99\%$ in the LBS report. The variance reported for the translation results is also lower, at $\approx 0.01451$ in contrast with the report's $\approx 0.06668$. Altogether, the Pearson correlation between the two values is $\approx 0.30356$, with the p-values for one and two tail both being lower than the alpha value chosen ($\alpha = 0.05$). This result enforces the fact that the translation exhibits a tendency towards male defaults, even when the female participation in that job is higher than expected.

@@

\section{Conclusions}

In this paper, we have provided preliminary evidence that statistical translation tools such as Google Translate can exhibit gender biases and a strong tendency towards male defaults. Although implicit, these biases possibly stem from the real world data which is used to train them, and in this context possibly provide a window into the way our society talks (and writes) about women in the workplace. In this paper, we suggest that and test the hypothesis that statistical translation tools can be probed to yield insights about stereotypical gender roles in our society -- or at least in their training data. By translating professional-related sentences such as \say{He/She is an engineer} from gender neutral languages such as Hungarian and Chinese into  English, we were able to collect statistics bout the asymmetry between female and male pronominal genders in the translation outputs. Our results show that male defaults are not only prominent but exaggerated in fields suggested to be troubled with gender stereotypes, such as STEM (Science, Technology, Engineering and Mathematics) jobs. And Because Google Translate typically uses English as a \emph{lingua franca}
% PEDRO: reference?
to translate between other languages (e.g. Chinese $\rightarrow$ English $\rightarrow$ Portuguese), our findings possibly extend to translations between gender neutral languages and non-gender neutral languages (apart from English) in general, although we have not tested this hypothesis.

Although not conclusive, our results seem to suggest that this phenomenon extends beyond the scope of the workplace, with the proportion of female pronouns varying significantly according to adjectives used to describe a person. Adjectives such as \emph{Shy} and \emph{Attractive} are predominantly translated with female pronouns, while \emph{Guilty} and \emph{Cruel} are almost exclusively translated with male pronouns. Different languages also seemingly have a significant impact in machine gender bias, with Hungarian exhibiting a better equilibrium between male and female pronouns than for example Chinese. Some languages such as Yoruba and Basque were found to translate sentences with gender neutral pronouns very often, although this is the exception rather than the rule and these languages also exhibit a high frequency of translation errors.

To solidify our results, we ran our pronominal gender translation statistics against the U.S. Bureau of Labor Statistics data on the frequency of women participation for each job position. Although Google Translate exhibits male defaults, this phenomenon may merely reflect the unequal distribution of male and female workers in some job positions. To test this hypothesis, we compared the distribution of female workers with the frequency of female translations, finding no correlation between said variables. Our data shows that Google Translate outputs fail to reflect the real-world distribution of female workers, under-estimating the expected frequency. That is to say that even if we do not expect a 50:50 distribution of translated gender pronouns, Google Translate exhibits male defaults in a greater frequency that job occupation data alone would suggest. The prominence of male defaults in Google Translate is therefore to the best of our knowledge yet lacking a clear justification.

We think this work can shed further light on what we believe is a pressing ethical difficulty arising from modern statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potential harmful effects of the pressing modern concerns of machine bias. We are optimistic that unbiased results can be obtained with little effort and marginal cost to the performance of current methods, to which current \emph{debiasing} algorithms in the scientific literature are a testament.
